Clinical advancement forecasting \thanks{\textit{\underline{Citation}}: \textbf{TBD}}

### 2024-04-09 paper/paper.tex
Start discussion section

\begin{abstract}

Selecting optimal drug targets for diseases represents a critical juncture in the creation of new medicines. In this study, we present an automated framework for selecting such targets through a synthesis of human, animal, and cell model evidence that features simple, highly interpretable statistical models. These models are optimized based on the historical clinical success of target-disease pairings, use only 33 predictors derived directly from individual lines of evidence in Open Targets and can predict clinical advancement for those pairings with a success rate that is 1.5-2x higher than heuristic, composite scores derived from the same evidence as well as common measures of genetic support that have already been established -- and observed in this study -- to confer a 2x higher likelihood of success. We find that this cumulative 3-4x advantage is maximized when also utilizing prior human clinical validation specific to targets and diseases independent of pairings, or other static properties of targets like conservation, expression specificity and gene essentiality. These predictors are rarely absent, which contrasts highly with the evidence for target-disease pairs that rarely exists among those that have ever reached clinical trials. The extent and non-random nature with which this absence manifests pose key challenges in the selection of performance measures and predictor derivation. In particular, we address the possibility for reverse causality between clinical outcomes and evidence discovery through the use of longitudinal predictors. This novel aspect of our work also enables us to directly evaluate which types of evidence are most likely to leak information about clinical outcomes when not temporalized. By formulating target selection as a machine learning problem in this manner, and after accounting for these unique challenges, this framework can be easily deployed to prioritize disease targets yet to undergo clinical development. We explore some of these present-day targets, with a focus on those that are tractable, and offer all code and data used in this analysis as a proof-of-concept for our more comprehensive, proprietary ranking platform.
% Note: OT multiplier (2x) taken from Sensitivity / Relative risk ratio distribution across configurations / mean@10
\end{abstract}

@@

\begin{itemize}
  \item Cover limitations with OT concerning temporalization for both evidence and drug approvals, and why only transitions from phase 2 are relevant for this work
  \item Discuss the possibility to do prospective evaluation with OT snapshots
  \item Discuss the limitations of the tractability analysis and why they might not be tractable targets
  \item Discuss why we present metrics like AP and ROC at all, and what kinds of use cases might maximize for something like ROC instead
  \item Expand on how the inclusion of time greatly improves tail performance and mention that this is consistent with its nature as a necessary but not sufficient condition for success, and it is likely that much of the value it adds in the tail of lower rankings could be captured and enhanced if other early indicators for the many reasons trials fail \cite{Razuvayevskaya2023.02.07.23285407} were also included (see more on this in Section~\ref{sec:discussion})
  \item We are focusing on prioritizing among targets/indications with genetic support, rather than expanding this space to find opportunities with weaker support
  \item "There is evidence that a 9.6\% vs. 13.8\% success rate for drugs from phase 1 trials to approval may mean a \$480 million difference in the median research and development cost required to bring a new drug to the market (Wouters, McKee, and Luyten 2020)." \cite{PMID:34930919}
  \item From \cite{PMID:33262371}: "It is important to bear in mind therefore that what we are measuring when looking at historical trial outcomes is not an unbiased measure of any given gene's true disease associations, but rather a view on how useful a given evidence source or analytical method has been for choosing drug targets based on current and historical drug discovery practices. Dramatic changes in these practices in the future could render some of our conclusions obsolete, though the fundamental observation that genetic association itself is retained in molecular networks will remain valid."
  \item Add select tractability and DepMap features?
\end{itemize}

>>>>

\begin{abstract}
Choosing which drug targets to pursue for a given disease is one of the most impactful decisions made in the global development of new medicines. This study examines the extent to which the outcomes of clinical trials can be predicted based on a small set of longitudinal (temporally labeled) evidence and properties of drug targets and diseases. We demonstrate a novel statistical learning framework utilizing both constrained and unconstrained linear and tree models, that can optimize predictions of a target-disease pair's likelihood of advancing beyond phase 2 clinical trials, outperforming existing baselines, including Open Targets composite scores, by 2x or greater, while preserving interpretability and counteracting common forms of bias. Utilizing a subset of our biomedical evidence base, non-negative linear models can produce simple weighting schemes across various types of human, animal, and cell model genomic, transcriptomic, proteomic, and clinical evidence to identify target-disease pairs with high probabilities of advancing beyond phase 2 trials, and uncover a range of previously undeveloped target-disease pairs poised for clinical success. In this study we further explore: i) how longitudinal treatment of evidence relates to leakage and reverse causality in biomedical research and how temporalized evidence can mitigate common forms of potential biases and inflation ii) the relative impact of different type of features on our predictions; and iii) an analysis of the space of currently undeveloped, tractable targets predicted with these methods to have the highest likelihood of clinical success.  To ease reproduction and deployment, none of our proprietary data is used outside of Open Targets and the described methods require no expert knowledge, and can support expansion of lines of evidence to further improve performance.
% Note: OT multiplier (1.5-2x) taken from Sensitivity / Relative risk ratio distribution across configurations / mean@10 and mean@20
\end{abstract}

@@

We have focused primarily on performance among the top 1\% of TD pairs throughout the majority of this study. Reasons for this are introduced in Section~\ref{sec:features} and Section~\ref{sec:metrics}. It is constructive to expand on this choice by comparing it to alternatives, and noting that it is quite likely that many other predictors could improve performance for the remaining 99\%. We offer some evidence of this here through the use of a feature indicating how long a TD pair has been in phase 2 trials, which improves wholistic measures of ranking performance like ROC by 12 points (see Figure~\ref{fig:performance_metric_mean_across_ta}).

\begin{itemize}
  \item Expand on how the inclusion of time greatly improves tail performance and mention that this is consistent with its nature as a necessary but not sufficient condition for success, and it is likely that much of the value it adds in the tail of lower rankings could be captured and enhanced if other early indicators for the many reasons trials fail \cite{Razuvayevskaya2023.02.07.23285407} were also included (see more on this in Section~\ref{sec:discussion})
  \item We are focusing on prioritizing among targets/indications with genetic support, rather than expanding this space to find opportunities with weaker support
  \item Discuss why we present metrics like AP and ROC at all, and what kinds of use cases might maximize for something like ROC instead
  \item Cover limitations with OT concerning temporalization for both evidence and drug approvals, and why only transitions from phase 2 are relevant for this work
  \item Discuss the possibility to do prospective evaluation with OT snapshots
  \item Discuss the limitations of the tractability analysis and why they might not be tractable targets
  \item "There is evidence that a 9.6\% vs. 13.8\% success rate for drugs from phase 1 trials to approval may mean a \$480 million difference in the median research and development cost required to bring a new drug to the market (Wouters, McKee, and Luyten 2020)." \cite{PMID:34930919}
  \item From \cite{PMID:33262371}: "It is important to bear in mind therefore that what we are measuring when looking at historical trial outcomes is not an unbiased measure of any given gene's true disease associations, but rather a view on how useful a given evidence source or analytical method has been for choosing drug targets based on current and historical drug discovery practices. Dramatic changes in these practices in the future could render some of our conclusions obsolete, though the fundamental observation that genetic association itself is retained in molecular networks will remain valid."
  \item Add select tractability and DepMap features?
\end{itemize}
